Reproduction frequency and offspring survival decline in elephant seals past prime age

### 2024-07-15 paper/paperv3.qmd
more small changes

# # Create POPULATION-LEVEL (i.e. without rand effects) predictions
# breed_pred_pop <- ggpredict(breed_age_mod, 
#                             terms = c("age10 [all]", "age_cat")) %>% 
#   as_tibble() %>% 
#   mutate(age = x * 10 + age_senesce,
#          age_cat = factor(group, levels = c("Young", "Old"))) %>% 
#   filter((age_cat == "Young" & age < age_senesce) |
#            (age_cat == "Old" & age >= age_senesce),
#          age <= 20) %>% 
#   left_join(n_breed, by = "age")

@@

(fig_selapp | fig_seldis) +
  plot_annotation(tag_levels = "A")
```

There was no evidence for selective appearance or disappearance influencing the relationship between maternal age and probability of reproduction. The selective appearance model (i.e., included a term for age-of-first-reproduction) had a similar AIC weight as the base model. However, in the selective appearance model, the coefficient for age at first reproduction was not significantly different than 0 (`r h1_firstrepro_95ci`). As a result, the predicted breeding probabilities for animals that first recruited at ages 4 and 7 (the 10th and 90th observed percentiles in the population) were not qualitatively different than the predictions of the base model [@fig-selective-h1]. Therefore the observed pattern is consistent with the senescence hypothesis (i.e., within-animal change with age).

@@

```{r}
#| label: selective-h2

h2_aic <- do.call(AIC, h2_selective_models) %>% 
  as_tibble() %>% 
  mutate(model = c("Base", 
                   "Selective appearance", 
                   "Selective disappearance", 
                   "Selective appearance and disappearance"),
         dAIC = AIC - min(AIC),
         rellik = exp(-0.5 * dAIC),
         weight = rellik / sum(rellik)) %>% 
  relocate(model)

# Selective appearance model, first repro 95% ci
h2_selapp_coef <- summary(h2_selective_models[[2]])$coef["firstrepro", ]
h2_firstrepro_95ci <- (h2_selapp_coef["Estimate"] + c(-1.96, 1.96) * h2_selapp_coef["Std. Error"]) %>% 
  make_95()

h2_aic

@@

fig_selapp | fig_seldis +
  plot_annotation(tag_levels = "A")
```

There was no evidence for selective appearance or disappearance influencing the relationship between maternal age and offspring survival. The selective appearance model (i.e., included a term for age-of-first-reproduction) had a slightly greater AIC weight than the base model. However, in the selective appearance model, the coefficient for age at first reproduction was not significantly different than 0 (`r h2_firstrepro_95ci`). As a result, the predicted breeding probabilities for animals that first recruited at ages 4 and 7 (the 10th and 90th observed percentiles in the population) were not qualitatively different than the predictions of the base model [@fig-selective-h2]. Therefore the observed pattern is consistent with the senescence hypothesis (i.e., within-animal change with age).

@@

First, we tested whether probability of detection was related to age. We found the proportion of unobserved animals increased with age. For this analysis, we excluded the first and last year of observation for each animal, because animals were necessarily observed in those years. For all years between, we determined whether animals were observed. Then we estimated the effect of age on probability of an animal being unobserved using a GLMM (binomial response, logit link) with random effects for year and animal ([@fig-unobserved]). The coefficient for age was `r missing_age_coef` (*p*\<0.0001).

@@

missing_years <- count(missing_seals, year_fct)
missing_pred <- expand_grid(age = 5:19, 
                            year_fct = factor(1989:2022)) %>% 
  mutate(age10 = (age - age_senesce) / 10) %>% 
  mutate(predicted = predict(missing_mod, 
                             newdata = missing_grid,
                             re.form = ~ (1 | year_fct)),
         se = predict(missing_mod, 
                      newdata = missing_grid,
                      se.fit = TRUE,
                      re.form = NA)$se.fit,
         conf.low = predicted - 1.96 * se,
         conf.high = predicted + 1.96 * se) %>% 
  left_join(missing_years, by = "year_fct") %>% 
  group_by(age) %>% 
  summarize(across(c(predicted, conf.low, conf.high), mean)) %>% 
  mutate(across(c(predicted, conf.low, conf.high), family(missing_mod)$linkinv))

missing_pred_years <- expand_grid(age = 5:19, 
                            year_fct = factor(1989:2022)) %>% 
  mutate(age10 = (age - age_senesce) / 10) %>% 
  mutate(predicted = predict(missing_mod, 
                             newdata = missing_grid,
                             type = "response",
                             re.form = ~ (1 | year_fct)))

missing_pred

missing_seals %>% 
  group_by(age) %>% 
  summarize(unobserved_mean = mean(unobserved),
            unobserved_se = sqrt(unobserved_mean * (1 - unobserved_mean) / n()),
            unobserved_lwr = unobserved_mean - unobserved_se,
            unobserved_upr = unobserved_mean + unobserved_se) %>% 
  ggplot(aes(age)) +
  geom_line(aes(y = predicted, group = year_fct), 
            missing_pred_years, 
            alpha = 0.2) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high),
              missing_pred,
              alpha = 0.2,
              fill = "#de77ae") +
  geom_line(aes(y = predicted),
            missing_pred,
            color = "#de77ae",
            linewidth = 2) +
  geom_pointrange(aes(y = unobserved_mean, 
                      ymin = unobserved_lwr, ymax = unobserved_upr)) +
  scale_x_continuous("Age", breaks = seq(5, 19, by = 4)) +
  scale_y_continuous("Unobserved proportion", 
                     labels = scales::label_percent(accuracy = 1),
                     limits = c(0, NA))
```

>>>>

@@

There was no evidence for selective appearance or disappearance influencing the relationship between maternal age and probability of reproduction. The selective appearance model (i.e., included a term for age-of-first-reproduction) had a similar AIC weight as the base model. However, in the selective appearance model, the coefficient for age at first reproduction was not significantly different than 0. As a result, the predicted breeding probabilities for animals that first recruited at ages 4 and 7 (the 10th and 90th observed percentiles in the population) were not qualitatively different than the predictions of the base model [@fig-selective-h1]. Therefore the observed pattern is consistent with the senescence hypothesis (i.e., within-animal change with age).

@@

@@

There was no evidence for selective appearance or disappearance influencing the relationship between maternal age and offspring survival. The selective appearance model (i.e., included a term for age-of-first-reproduction) had a slightly greater AIC weight than the base model. However, in the selective appearance model, the coefficient for age at first reproduction was not significantly different than 0. As a result, the predicted breeding probabilities for animals that first recruited at ages 4 and 7 (the 10th and 90th observed percentiles in the population) were not qualitatively different than the predictions of the base model [@fig-selective-h2]. Therefore the observed pattern is consistent with the senescence hypothesis (i.e., within-animal change with age).

@@

First, we tested whether probability of detection was related to age. We found the proportion of unobserved animals increased with age. For this analysis, we excluded the first and last year of observation for each animal, because animals were necessarily observed in those years. For all years between, we determined whether animals were observed. Then we estimated the effect of age on probability of an animal being unobserved using a GLMM (binomial response, logit link) with random effects for year and animal ([@fig-unobserved]). The coefficient for age was r missing_age_coef (*p*\<0.0001).

@@

missing_years <- count(missing_seals, year_fct)
missing_pred <- expand_grid(age = 5:19, 
                            year_fct = factor(1989:2022)) %>% 
  mutate(age10 = (age - age_senesce) / 10) %>% 
  mutate(predicted = predict(missing_mod, 
                             newdata = .,
                             re.form = ~ (1 | year_fct)),
         se = predict(missing_mod, 
                      newdata = .,
                      se.fit = TRUE,
                      re.form = NA)$se.fit,
         conf.low = predicted - 1.96 * se,
         conf.high = predicted + 1.96 * se) %>% 
  left_join(missing_years, by = "year_fct") %>% 
  group_by(age) %>% 
  summarize(across(c(predicted, conf.low, conf.high), mean)) %>% 
  mutate(across(c(predicted, conf.low, conf.high), family(missing_mod)$linkinv))

missing_pred_years <- expand_grid(age = 5:19, 
                            year_fct = factor(1989:2022)) %>% 
  mutate(age10 = (age - age_senesce) / 10) %>% 
  mutate(predicted = predict(missing_mod, 
                             newdata = .,
                             type = "response",
                             re.form = ~ (1 | year_fct)))

missing_pred_pop <- tibble(age = 5:19,
                           age10 = (age - age_senesce) / 10) %>% 
  mutate(predicted = predict(missing_mod, 
                             newdata = .,
                             type = "response",
                             re.form = NA))

missing_seals %>% 
  group_by(age) %>% 
  summarize(unobserved_mean = mean(unobserved),
            unobserved_se = sqrt(unobserved_mean * (1 - unobserved_mean) / n()),
            unobserved_lwr = unobserved_mean - unobserved_se,
            unobserved_upr = unobserved_mean + unobserved_se) %>% 
  ggplot(aes(age)) +
  geom_line(aes(y = predicted, group = year_fct), 
            missing_pred_years, 
            alpha = 0.2) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high),
              missing_pred,
              alpha = 0.2,
              fill = "#de77ae") +
  geom_line(aes(y = predicted),
            missing_pred,
            color = "#de77ae",
            linewidth = 2) +
  geom_line(aes(y = predicted),
            missing_pred_pop,
            linewidth = 1.5,
            linetype = "dotted",
            alpha = 0.5) +
  geom_pointrange(aes(y = unobserved_mean, 
                      ymin = unobserved_lwr, ymax = unobserved_upr)) +
  scale_x_continuous("Age", breaks = seq(5, 19, by = 4)) +
  scale_y_continuous("Unobserved proportion", 
                     labels = scales::label_percent(accuracy = 1),
                     limits = c(0, NA))
```
